analysis_uown

### 2021-11-12 code/analysis_code/analysis_uown.Rmd
LASSO_Model - CC

### Required packages
```{r}
library(tidyverse)
library(tidymodels)
library(here)
```

@@

>>>>

### Required packages
```{r}
library(tidyverse) #Working with multiple Tidy packages
library(tidymodels) #Building models
library(dplyr) #data manipulation
library(here) #setting pathways for saving files
library(rpart) #Model fitting
library(ranger) #Model fitting
library(glmnet) #Model fitting
```

@@

# Part 3: LASSO Modeling for Ideal Combination of Stream Health Indicators to Predict Biological SCore and E. Coli Colony Forming Units in MIDO Data.

# Biological Score

## Data Set Up:

_Take ID and time variables out of MIDO Data frame. This will help reduce errors when we try to run the Model._
```{r}
MIDO_LASSO <- MIDO %>%
  select(-"WSID", -"stream_ID", -"month", -"year", -"day", -"pH")
```

_log transform E. coli data to fix bias in E. coli data_
```{r}
MIDO_LASSO <- MIDO_LASSO %>%
  mutate(e.coli.cfu = log10(MIDO_LASSO$e.coli.cfu))
```

_Set seed: This sets a random number generator with initial (pseudo)random values set as "123". We will need a series of random numbers created for our machine learning analysis._
```{r}
#Set random number generator.
set.seed(123)
```

 _Test/Train data split: Split data into a training data set and a test data set. The training set will contain most of the original data (70%), while the test data set will have a smaller portion (30%). The training data is used to fit a model and the test data is to assess how good of a fit the data is._

## Null Model For MIDO Data

_5-fold cross validation, 5 times repeated for MIDO data: Here we are setting a cross-validation of the machine learning models. Cross-validation is used to measure how the results of our machine learning models will generalize to an independent data set. As such, the folds created will be be 5 random sub-samples of the train data set to test the validity of our models within the train data set. The 5x5 structure is arbitrary._
```{r}
#Create folds for LASSO Cross Validation.
fold_BS <- vfold_cv(MIDO_LASSO, v = 5, repeats = 5, strata = biological_score)
```

_Creating the recipe for Biological Score vs all predictors_

```{r}
BS.recipe <- 
  recipe(biological_score ~ ., data = MIDO_LASSO) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_predictors())

BS.recipe
```
_Setting linear regression model to assess relationship between Biological Score (outcome) and all other predictor variables._
```{r}
lm_mod <- linear_reg() %>% set_engine("lm") %>% set_mode("regression")
```

_However, first we need to create our null model to test against._

### Null Model:

_Creates null model recipe. When we call this term, it will indicate in our workflow that Biological Score will be predicted by a value of 1 (NULL)._ 
```{r}
Null_recipe_lm <- recipe(biological_score ~ 1, data = MIDO_LASSO)
```

_Creating the Workflow: this creates a set workflow for running a null linear regression model with body temperature as the outcome._
```{r}
null_wf <- workflow() %>% add_model(lm_mod) %>% add_recipe(Null_recipe_lm)
```

_Here, I am going to fit the null model created in the above workflow to the folds made from the train data set._
```{r}
null_lm <- fit_resamples(null_wf, resamples = fold_BS)
```
_Calculate RMSE for the train data linear model._
```{r}
Null_Met <- collect_metrics(null_lm)

Null_Met
```
_RMSE = 6.63, with a standard deviation of 0.13. This will serve as our check to test our models against latter on.

## LASSO Model

_Specifying The Model: LASSO_
```{r}
lasso_mod <- 
  linear_reg(penalty = tune(), mixture = 1) %>% 
  set_engine("glmnet")
```

_Creating a Workflow: LASSO_
```{r}
lasso_wf <- workflow() %>%
  add_model(lasso_mod) %>%
  add_recipe(BS.recipe)
```

_Create Tuning Grid: LASSO_
```{r}
LASSO_grid <- grid_regular(penalty(), levels = 50)
```

_Cross Validation with tune_grid(): LASSO_
```{r}
lasso_resample <- 
  lasso_wf %>%
  tune_grid(resamples = fold_BS,
            grid = LASSO_grid,
            control = control_grid(verbose = FALSE, save_pred = TRUE),
            metrics = metric_set(rmse))

lasso_resample %>%
  collect_metrics()
```

_Plot model performance using autoplot()_
```{r}
#Plot of MIDO data LASSO model performance
lasso_resample %>%
  autoplot()
```

_Showing and selecting best performing Models_
```{r}
#Showing best performing LASSO models
lasso_resample %>%
  show_best()

#Selects best performing model
best_lasso <- lasso_resample %>%
  select_best()
```

_This shows that model 47 is the best performing models (RMSE = 6.50; STE = 0.16). However, it doesn't really perform any better than the null model, making it a bad fit to the data._

_Creating final fit based on best model permutation and plotting predicted values from that final fit model_
```{r}
lasso_final_wf <- 
  lasso_wf %>% 
  finalize_workflow(best_lasso)

lasso_final_wf

#Create workflow for fitting model to train_data2 predictions
lasso_final_fit <- 
  lasso_final_wf %>%
  fit(MIDO_LASSO) 
```

_Calculating residuals:_
```{r}
#Manually calculate residuals for MIDO data LASSO models between real and predicted values.
lasso_residuals <- lasso_final_fit %>%
  augment(MIDO_LASSO) %>% #use augment() to make predictions from train data
  select(c(.pred, biological_score)) %>%
  mutate(.resid = biological_score - .pred) #calculate residuals and make new row.

lasso_residuals
```
_model predictions from tuned model vs actual outcomes_
```{r}
lasso_pred_plot <- ggplot(lasso_residuals, 
                          aes(x = biological_score, 
                              y = .pred)) + 
  geom_point() + 
  labs(title = "Predictions vs Actual: LASSO", 
       x = "Biological Score Outcome", 
       y = "Biological Score Prediction")
lasso_pred_plot
```
_plot residuals vs predictions_
```{r}
lasso_residual_plot <- ggplot(lasso_residuals, 
                              aes(y = .resid, 
                                  x = .pred)) + 
  geom_point() + 
  labs(title = "Predictions vs Residuals: LASSO", 
       x = "Biological Score Prediction", 
       y = "Residuals")
plot(lasso_residual_plot) #view plot
```

# LASSO: E. coli cfu

## Data Set Up:

_Set seed: This sets a random number generator with initial (pseudo)random values set as "123". We will need a series of random numbers created for our machine learning analysis._
```{r}
#Set random number generator.
set.seed(123)
```

 _Test/Train data split: Split data into a training data set and a test data set. The training set will contain most of the original data (70%), while the test data set will have a smaller portion (30%). The training data is used to fit a model and the test data is to assess how good of a fit the data is._

## Null Model For MIDO Data

_5-fold cross validation, 5 times repeated for MIDO data: Here we are setting a cross-validation of the machine learning models. Cross-validation is used to measure how the results of our machine learning models will generalize to an independent data set. As such, the folds created will be be 5 random sub-samples of the train data set to test the validity of our models within the train data set. The 5x5 structure is arbitrary._
```{r}
#Create folds for LASSO Cross Validation.
fold_EC <- vfold_cv(MIDO_LASSO, v = 5, repeats = 5, strata = e.coli.cfu)
```

_Creating the recipe for Biological Score vs all predictors_

```{r}
EC.recipe <- 
  recipe(e.coli.cfu ~ ., data = MIDO_LASSO) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_predictors())

EC.recipe
```
_Setting linear regression model to assess relationship between Biological Score (outcome) and all other predictor variables._
```{r}
lm_mod <- linear_reg() %>% set_engine("lm") %>% set_mode("regression")
```

_However, first we need to create our null model to test against._

### Null Model:

_Creates null model recipe. When we call this term, it will indicate in our workflow that E. Coli cfu will be predicted by a value of 1 (NULL)._ 
```{r}
Null_recipe_lm_EC <- recipe(e.coli.cfu ~ 1, data = MIDO_LASSO)
```

_Creating the Workflow: this creates a set workflow for running a null linear regression model with E. coli cfu as the outcome._
```{r}
null_wf_EC <- workflow() %>% add_model(lm_mod) %>% add_recipe(Null_recipe_lm_EC)
```

_Here, I am going to fit the null model created in the above workflow to the folds made from the train data set._
```{r}
null_lm_EC <- fit_resamples(null_wf_EC, resamples = fold_EC)
```

_Calculate RMSE for the train data linear model._
```{r}
Null_Met_EC <- collect_metrics(null_lm_EC)

Null_Met_EC
```
_RMSE = 0.52, with a standard deviation of 0.014. This will serve as our check to test our models against latter on._

## LASSO Model

_Specifying The Model: LASSO_
```{r}
lasso_mod <- 
  linear_reg(penalty = tune(), mixture = 1) %>% 
  set_engine("glmnet")
```

_Creating a Workflow: LASSO_
```{r}
lasso_wf_EC <- workflow() %>%
  add_model(lasso_mod) %>%
  add_recipe(EC.recipe)
```

_Create Tuning Grid: LASSO_
```{r}
LASSO_grid <- grid_regular(penalty(), levels = 50)
```

_Cross Validation with tune_grid(): LASSO_
```{r}
lasso_resample_EC <- 
  lasso_wf_EC %>%
  tune_grid(resamples = fold_EC,
            grid = LASSO_grid,
            control = control_grid(verbose = FALSE, save_pred = TRUE),
            metrics = metric_set(rmse))

lasso_resample_EC %>%
  collect_metrics()
```

_Plot model performance using autoplot()_
```{r}
#Plot of MIDO data LASSO model performance
lasso_resample_EC %>%
  autoplot()
```

_Showing and selecting best performing Models_
```{r}
#Showing best performing LASSO models
lasso_resample_EC %>%
  show_best()

#Selects best performing model
best_lasso_EC <- lasso_resample_EC %>%
  select_best()
```

_This shows that model 41 is the best performing models (RMSE = 0.50; STE = 0.012). However, it doesn't really perform any better than the null model, making it a bad fit to the data._

_Creating final fit based on best model permutation and plotting predicted values from that final fit model_
```{r}
lasso_final_wf_EC <- 
  lasso_wf_EC %>% 
  finalize_workflow(best_lasso_EC)

lasso_final_wf_EC

#Create workflow for fitting model to train_data2 predictions
lasso_final_fit_EC <- 
  lasso_final_wf_EC %>%
  fit(MIDO_LASSO) 
```

_Calculating residuals:_
```{r}
#Manually calculate residuals for MIDO data LASSO models between real and predicted values.
lasso_residuals_EC <- lasso_final_fit_EC %>%
  augment(MIDO_LASSO) %>% #use augment() to make predictions from train data
  select(c(.pred, e.coli.cfu)) %>%
  mutate(.resid = e.coli.cfu - .pred) #calculate residuals and make new row.

lasso_residuals_EC
```
_model predictions from tuned model vs actual outcomes_
```{r}
lasso_pred_plot_EC <- ggplot(lasso_residuals_EC, 
                          aes(x = e.coli.cfu, 
                              y = .pred)) + 
  geom_point() + 
  labs(title = "Predictions vs Actual: LASSO", 
       x = "E. coli cfu Outcome", 
       y = "E. coli cfu Prediction")
lasso_pred_plot_EC
```
_plot residuals vs predictions_
```{r}
lasso_residual_plot_EC <- ggplot(lasso_residuals_EC, 
                              aes(y = .resid, 
                                  x = .pred)) + 
  geom_point() + 
  labs(title = "Predictions vs Residuals: LASSO", 
       x = "E. coli cfu Prediction", 
       y = "Residuals")
plot(lasso_residual_plot_EC) #view plot
```
